Ensemble Transport Adaptive Importance Sampling

### 2019-01-04 response_JUQ_2.tex

\vspace{0.3cm}{\bf Answer:} #2}

\begin{document}
\title{Response to Referee Comments: Parallel Adaptive Importance Sampling}
\maketitle
First of all, we would like to thank the referees and the associate editor for
their time in reading and commenting on our manuscript. In this document
we will address each of the points raised, and describe any relevant
changes that have been made to the paper. We have made several substantial changes, including the renaming of the algorithm and paper from ``parallel adaptive importance sampling'' (PAIS) to ``ensemble transport adaptive importance sampling'' (ETAIS). We have also replaced three of the old numerical examples with two ones, one a bimodal example, and the second a more challenging problem for Bayesian inversion of the Lorenz '63 equations. We believe that the paper is stronger, and thank the reviewers and associate editor for their input.

\section*{Referee 1 comments}

\comment{This is supposed to be a parallel algorithm and yet it's implemention is serial. Furthermore, the authors don't seem to be entirely sure that parallelization will work since they comment that it's effectiveness will depend on various factors. The title seems somehow misleading to me. Maybe if the "parallel" part of the title is removed and the discussion about parallelization is treated as a possibility instead of a critical part of the algorithm, then the document will be less confusing.}{Despite the fact that we felt that we had demonstrated the feasibility of the parallelisation of this method, we agree with the referee that we have not demonstrated this in the numerics. Furthermore, this algorithm delivers a speed-up even when run in serial. As such, we are happy to change the name of the algorithm to ``Ensemble Transport Adaptive Importance Sampling'' (ETAIS). This required some rewriting throughout the paper. In particular, we have toned down the emphasis on parallelisation, and put more emphasis on the fact that this algorithm also delivers speed-up for serial implementaions. See the second paragraph of the discussion.}

\comment{Also, the comparison between serial PAIS and the naive parallel MCMC is strange if there is no parallelization.}{The naive parallel chains are run for long enough that they are reasonably similar to one long chain, and as such we didn't feel that it would be worthwhile to rerun all of these experiments.}

\comment{It will be interesting that the authors show the PAIS using AMS with the examples. They show the difference between the AMS and the ETPF but not within the PAIS.}{We assume that the referee meant the AMR? As the AE asked for a new example to replace the second example, we took this as an opportunity to compare ETAIS with the ETPF, AMR and bootstrap resamplers, which is now presented in the new first example. The numerics demonstrate that superior samplers lead to more stable and accurate sampling, with ETPF algorithm narrowly outperforming the AMR algorithm, with the bootstrap version following up in the rear. This helps support our claim that state-of-the-art resampling methods can lead to more efficient AIS algorithms.}

\comment{There is an error at the end of page 24.}{Many thanks for spotting the missing parenthesis, this has now been corrected.}

@@

\section*{Referee 2 comments}
\comment{The paper is ready for publication.}{Many thanks. We hope you agree that the additional changes have further improved the paper.}

\section*{Associate Editor comments}

\comment{The authors have made a substantial improvement over previous version of the paper. However, there are a number of points that I would like to see clarified before the paper is accepted for publication.\\
1) The posterior sometimes is called $\mu$ sometimes $\pi$ some times $\pi(x|D)$ etc. In fact, there is a confusion in Alg. 1 in p.6 with this.}{Thanks for noticing this. We have carefully gone through the paper and removed all uses of $\mu$/measures in reference to distributions, and used only $\pi$ for posterior density throughout.}

\comment{2) Choosing the proposal (kernel) in a MH is all the main and commonly the only problem in MCMC. PAIS also needs a proposal of exactly the same kind. Why can we expect that within PAIS this will be easier? An indeed, PAIS will perform better that a naively parallelized MH, both using the same proposal. How PAIS saves some of the problems in finding good proposals?}{As discussed in the paper, there are many possible choices for the proposal kernel around each of the ensemble members. The only thing that we require to ensure convergence of an AIS algorithm, is the absolute continuity of the kernel with respect to the posterior. However, to ensure stability of the algorithm, we also require that the tails of the proposal distributions are at least as fat as the target. In practice, this means that picking kernels which are of the same type as the prior is usually a good idea. The ETAIS methodology does not, however, greatly benefit from the use of more informed and expensive proposals, such as are seen in the standard M-H framework, where gradient information can be used to improve proposals (MALA, HMC, etc), since the information about the whole target is represented by the ensemble. As mentioned in the paper, we have conducted similar studies with MALA proposals, and although we sometimes observe a slightly quicker burn-in period, overall the performance is largely the similar to that of a RW proposal, at twice the cost. There was already a discussion on this at the end of Section 3, but we have added some more detail.}

\comment{3) It might be trivially to do so, but the authors only briefly mention a prove for the convergence of their algorithm. In fact, is it ok to use a dependent proposal in importance sampling? ... no regularity conditions are needed? Why Alg. 3 works? resampling "introduces error" ... is this relevant? does the error cascades down in each iteration? In the examples the PAIS seem to work ok, but, do we need any regularity condition? Please make your convergence claims more explicit and take more time to explain them. In particular, do we have the correct invariant distribution?}{This algorithm belongs to a family of methods for which convergence has been proven in the literature, and as such, we have referenced this proof, and as such we did not feel it appropriate to provide further details. One detail was missing from our statement, which is that the posterior distribution must be absolutely continuous with respect to the proposal distribution, which we have now added.}

\comment{4) There are too many examples. The univariate gaussian is not very informative and the univariate bimodal, is not either since both modes are of the same scale. I would rather see these two examples turned into one single one with a 2D mixture of two separated gaussians with contrasting scales (variance covariance matrices).}{We agree that the numerical experiments section is very long. However the examples presented do each demonstrate a different aspect of the advantages of our approach. We have taken this comment on board, however, and have drastically reduced the length of example 1, in which we demonstrate that tuning the proposal variances using the effective sample size is approximately equivalent to using the $L^2$ errors in the target, which is only available to us as we have picked a trivial target. We have replaced the second bimodal example, with a bimodal example as requested by the AE with differing scales in the covariances of the two modes. We have also taken this opportunity to demonstrate the differences in results when using the different resamplers discussed in the paper. We have also included a comparison with a standard bootstrap resampler. The new numerics demonstrate the improvement in stability and convergence of the method with the use of more advanced resampling methods. The mixture model example demonstrates the algorithm working in a higher dimensional multimodal setting. The final new example (see later comments) demonstrates the speed-ups achieved using our approach for a challenging problem with an expensive likelihood, requiring the approximation of a differential equation. We believe that each example serves a purpose, but we have also reduced the size of this section by 3-4 pages.}

@@

\comment{5) In the examples the authors include very ad hoc strategies like the 'scout' chain and specially the very ad hoc strategy described in section 7.4.2 (!). The authors, in my opinion, do not really discuss the implementation of the PAIS algorithm but a case-by-case ad hoc version of it. This is clearly unacceptable.}{The scout chains were only used in the old bimodal example, and have since been cut. This approach has already been used in the pMC literature to stabilise the importance sampling by fattening the tails of the proposal distribution, but we have no problem with removing it from this paper.

  As for the proposal kernels used in 7.4.2, we disagree that this approach is ``ad-hoc''. These are chosen simply because the proposal distribution must only be supported by [0,1]. This relates to the previous answers we have given to comments made by the referee above. We have gone into more details regarding this and explaining the choices.}

\comment{6) Please clarify the first phrase of section 7.4.2, which, as far as I see, is simply twice wrong. There is a (very simple!) analytic version of the posterior AND with the right proposal MH can sample from it (!).}{We have removed this phrase from the paper.}
% {We have amended the statement about the analytic form of the posterior. We are not aware of any M-H methods which can reliably sample from multimodal densities with well-separated modes which are not known a priori. HMC trajectories can be hand-tuned to make transitions between modes, but only with information regarding the separation. Ensemble-based methods appear to us to be a good candidate for reliable sampling from such distributions without such a hand-tuning. However, we agree to remove the claim.}

\comment{7) Although the chaotic Lorenz 93 ODE is used, it is used in such a short time scale that the posterior is quite well behaved (unimodal and fairly gaussian and only some correlation is seen between to parameters). Since the priors are normal, this regime suggest a linear part of the FM. The authors should attempt something more challenging, by running the system into a highly nonlinear regime. Otherwise, why consider this example?}{The reason for considering the example was that the AE had previously asked us to produce an example in the regime where we suggest that we get the biggest win using our proposed method, i.e. those where the likelihood evaluations have significant cost due to requirement of the solution of a differential equation. In this example, we have demonstrated a significant speed-up over the equivalent M-H algorithm. We do understand however the AE's desire for a more challenging inverse problem in itself, and so we have created a new example which has a prior distribution with larger variance. The push-forward of this prior distribution is multimodal on the timescale of the observations (see the figure below), showing that trajectories do pass close to a hyperbolic point in the assimilation window. This also results in a longer thinner posterior, which as can be seen from the numerics, RWMH really struggles to sample from efficiently. The ETAIS implementation handles this problem with ease, for a big speed-up. The posterior is not multimodal, but it is even thinner/more elongated than previously, and thus more challenging Chaotic behaviour will in general lead to only a very small area of the state space being consistent with the observations, and so often (as in this case), the posterior is challenging because it is very thin in certain directions. Overall, the example demonstrates the speed-up of our approach for a problem where the likelihood is expensive.}

\begin{figure}[htp]
\includegraphics[width=0.3\textwidth]{figures/xhist}
\includegraphics[width=0.3\textwidth]{figures/yhist}
\includegraphics[width=0.3\textwidth]{figures/zhist}
\caption{Histograms from 1e7 Monte Carlo runs of the push forward of the new prior for the Lorenz '63 problem}
  \end{figure}

  \comment{8) Do not use an explicit Euler solver, use any higher order, off the shelve solver.}{Many of the off-the-shelf higher order solvers are just as inappropriate to use for the approximation of the solution of the Lorenz '63 equations. We agree that the explicit Euler method with a small timestep is not an optimal choice, but it is a simple choice, which still leads to a chaotic nonlinear system. We argue that it is sufficient for the purpose of comparing the efficiency of two different Monte Carlo methods, both of which are using the same approximation of the likelihood.}

  %{Many of the off-the-shelf higher order solvers are just as inappropriate to use for the approximation of the solution of the Lorenz '63 equations, since they are chaotic, and as such \emph{any} numerical error is quickly amplified through the dynamics. Moreover, the focus of this paper is not on these applications, but in the statistical algorithm. In this example, we are simply looking for a non-linear mapping such that we can demonstrate the speed-up that is gained from using ETAIS as opposed to standard M-H sampling, and we believe that a simple Euler approximation provides such a mapping.}
\end{document}

### 2019-01-04 ETAIS.tex
A lot of changes
Did most of the changes. Things still to do: The plots and changes to
the final example.

\subsection{Bayesian inverse problems}

In this paper, we focus on the use of MCMC methods for characterising
posterior probability distributions arising from Bayesian inverse problems. We
wish to learn about a particular unknown quantity $x$, of which we are
able to make direct or indirect noisy observations. For now
we say that $x$ is a member of a Hilbert
space $X$.

The quantity $x$ is mapped on to observable space by the observation
operator $\mathcal{G}:X \to\mathbb{R}^d$. We assume that the
observations, $\data$, are subject to Gaussian noise,
\begin{equation}\label{eqn:obs}
	\data = \mathcal{G}(x) + \varepsilon, \qquad \varepsilon \sim \mu_{\varepsilon} = \mathcal{N}(0,\Sigma).
\end{equation}

@@

Then, according to Bayes' theorem, the posterior distribution that we
are interested in characterising is given, up to a constant of proportionality, by
\begin{equation*}
	\pi_Y \propto \exp \left ( -\Phi(x^*) \right )\pi_0(x^*),
\end{equation*}
where $\pi_0$ is the prior density on the quantity of interest $x^*$.

@@

In this work we use the resampler used within these particle filters. We
approximate the posterior distribution, $\mu_Y$, with a sample
of weighted particles, $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
the weights would be found by incorporating new observed data whereas
here we simply use importance weights. This distribution, $\hat{\mu}_Y
\approx \mu_Y$, can then be resampled, using the ETPF or otherwise,
into a new distribution, $\hat{\eta} \approx \mu_Y$,
\begin{equation}\label{eqn:resampler}
	\hat{\mu}_Y = \sum\limits_{i=1}^M \! w_i\delta_{y_i}(\cdot)
	 \quad \xrightarrow{\text{ETPF}} \quad \sum\limits_{i=1}^M \!
	 \frac{1}{M} \delta_{x_i}(\cdot) = \hat{\eta}.
\end{equation}This
new sample $\{x_i\}_{i=1}^M$ is equally weighted.

@@

We wish to sample states $x \in X$ from a posterior
probability distribution $\mu_D$, where $D$ represents our data which
we wish to use to infer $x$. Since we have $M$ ensemble members, we
represent the current state of all of the Markov chains as a vector
$\X = [x_1,x_2,\ldots,x_M]^\top$. We are also given a transition kernel
$\nu(\cdot,\cdot)$, which might come from an MCMC method, for example
the random walk Metropolis-Hastings proposal density $\nu(\cdot,x) \sim
\mathcal{N}(x,\beta^2)$, where $\beta^2\in \mathbb{R}$ defines the
variance of the proposal.

@@

Care needs to be taken when choosing kernel(s) for the proposal distribution to
ensure that the overall mixture is absolutely continuous with respect to
the posterior distribution. If this is not the case, importance
weights can become high variance where samples are made away from
modes in the distribution. This will
drastically decrease the efficiency of sampling.

@@

\section{Consistency of ETAIS}\label{sec:consistency}

@@

\section{A useful property of the ETAIS algorithm for multimodal distributions}

@@

\subsection{Effect of resampler quality}
\label{sec:bimodal}
{\red
In this second example we investigate the behaviour of the ETAIS
algorithm when applied to a bimodal problem, given by a mixture of two
Gaussians,
\[\pi(x) = 0.2\pi_1(x) + 0.8\pi_2(x),\]
where $\pi_1$ is the density of a $\mathcal{N}\left (\begin{pmatrix} 1
    \\ 1 \end{pmatrix}, \begin{pmatrix} 0.1 & 0 
    \\ 0 & 0.1 \end{pmatrix} \right )$ random variable and $\pi_2$ is the
  density of a $\mathcal{N}\left (\begin{pmatrix} -5
    \\ -5 \end{pmatrix}, \begin{pmatrix} 2.75 & -2.25 
    \\ -2.25 & 2.75 \end{pmatrix} \right )$ random variable.}

@@

Figure \ref{kshdbf} shows the bimodal target density, and convergence
plots for the RWMH and ETAIS-RW with ETPF, AMR and bootstrap
resamplers. As expected, a higher quality resampler leads to better
proposal distributions, which in turn leads to greater stability and
more reliable convergence. However, this also shows that the AMR is a
good greedy approximation of the ETPF, and for a fraction of the cost
when the number of ensemble members is larger, as shown in Section
\ref{sec:AMR}. The RWMH algorithm fails to converge efficiently since chains only very rarely
make switches between the two modes, leading to very slow mixing. This
demonstrates a big advantage of ensemble-based methods when the target
is multimodal.

@@

\subsubsection{Implementation}
\label{sec:mixture_implementation}

We have no analytic form for the posterior distribution for this
model, and MH cannot reliably sample from this type of bimodal
distribution. We do however know that the probability density should
be evenly divided between the two modes. This is due to the symmetric
prior for $p$, and and the same priors being assigned to $\mu_1$ and
$\mu_2$, and also to $\sigma^2_1$ and $\sigma^2_2$. To decide which
mode a sample belongs to we define a plane which bisects the posterior
so that each point on this plane lies exactly halfway between the two
true solutions to the inverse problem i.e. the value of $\theta$ used
to generate the data, and also the $\theta$ obtained by a relabelling
of the parameters. Now that we can assign a sample to a particular
mode, we can calculate the density in each mode by summing the weights
associated to all samples in that mode,
\[
	\bar{w}_k = \sum\limits_{i=1}^N w_iI_{X_i \in \text{Mode $k$}}, \quad k = 1, 2,
\]
and the relative error in the amount of density in each mode is then
\begin{equation}\label{eqn:mode_prop}
	w_\text{error} = 2\left|\frac{\bar{w}_1}{\bar{w}_1+\bar{w}_2} - \frac{1}{2}\right|.
\end{equation}

Since the probability $p$ is constrained to lie in the interval
$[0,1]$, and the variances must be positive, it can be wasteful to use
Gaussian proposal distributions. For this example we make proposals by
matching the proposal distribution to the prior, scaling the variance
and centring the proposal at the previous value. So
\[
	q_p \sim \text{Beta}(\delta^{-2}p, \delta^{-2}(1-p)), \quad q_{\mu_{1,2}} \sim \mathcal{N}(\mu_{1,2}, 4\delta^2) \quad \text{and} \quad q_{\sigma^2_{1,2}} \sim \text{Gamma}(\alpha^*, \beta^*),
\]
where $\alpha^* = \sigma^2_{1,2}\beta^*$, $\beta^* =
\sigma^2_{1,2}/2\delta^2$ and $\delta$ is a scaling parameter to be
tuned. This means that our proposal distributions will not be a
mixture of multivariate Gaussians, but independent mixtures of
univariate Beta, Gamma and Gaussian distributions.

@@

For this example, we observe noisily the position of a particle at ten
equally spaced points in the time interval $t \in (0, 1]$. The Lorenz
equations are solved with the chaotic parameters given above and the
initial condition $\mathbf{x}_0^1$. A time step of $h=1\times 10^{-3}$
was used to evolve the equations numerically with the explicit Euler
method. The noise added to each observation was taken from the
distribution $\mathcal{N}(0, 0.1^2I$. Priors for the initial condition
coordinates were given by
\[
	x_0 \sim \mathcal{N}(-0.5, 0.03^2), \quad y_0 \sim\mathcal{N}(-0.5, 0.03^2) \quad \text{and} \quad z_0 \sim \mathcal{N}(15, 0.03^2).
\]
The posterior density function takes the form
\[
	\pi(\mathbf{x}_0|D) \propto \exp\left\{-\frac{1}{2}\|\mathcal{G}(\mathbf{x}_0) - D\|^2_\Sigma\right\}\mathcal{N}(\mathbf{x}_0; m, 0.03^2\text{I}),
\]
where $\mathcal{G}$ calculates the solution to the Lorenz equations, using the Euler method with a time step of $h=1\times 10^{-3}$ starting at the initial condition $\mathbf{x}_0$, and returns the position at $t = 0.1, ..., 1$. The mean of the prior distribution $m = [-0.5, -0.5, 15]$.

\begin{figure}[htb]
\centering
\includegraphics[width=0.8\textwidth]{"figures/Lorenz_posterior"}
\caption{Posterior distribution for the initial condition of the particle with motion governed by the Lorenz equations as described in Section~\ref{sec:lorenz_target}.}
\label{fig:Lorenz_posterior}
\end{figure}

@@

Both the MH and ETAIS algorithms use an ensemble size of $M=500$. The ETAIS algorithm uses the AMR resampling algorithm. The optimal scaling parameters are calculated for both algorithms using test runs which are not included in the convergence cost calculations. Convergence graphs are produced using 32 repeats of each algorithm with each simulation producing one million samples.

@@

The optimal scaling parameters for the MH and ETAIS algorithms are given in Table~\ref{table:Lorenz_opt_parameters}.

\begin{table}[!htb]
      \centering
        \begin{tabular}{|l|r|r|}
	\hline
	Algorithm	& MH & ETAIS \\ \hline
	$\delta^*$	& 1.0e-1     & 1.3e-2 \\
	\hline
	\end{tabular}
	\vspace{1mm}
	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the ETAIS algorithm is optimised using the effective sample size.}
	\label{table:Lorenz_opt_parameters}
\end{table}

\begin{figure}[htb]
\centering
\includegraphics[width=0.8\textwidth]{"figures/Lorenz_convergence"}
\caption{Convergence of the sample mean to the posterior mean for the MH and ETAIS algorithms. Posterior distribution as described in Section~\ref{sec:lorenz_target}.}
\label{fig:Lorenz_convergence}
\end{figure}

@@

\red{The implementations of the ETAIS that have been presented in this
  paper, have been run in serial, but we argue that, particularly for
  inverse problems with expensive likelihoods, for example in the case
  where this requires the approximation of the solution of a
  differential equation, that this method is an excellent candidate
  for parallelisation, with the likelihood evaluations for each
  particle being computed on a different processor. This said, even in
  parallel, the numerical results that we have presented demonstrate
  that even a serial implementation can outperform standard
  Metropolis-Hastings methods.}

>>>>

\subsection{Bayesian inverse problems}
In this paper, we focus on the use of MCMC methods for characterising
posterior probability distributions arising from Bayesian inverse problems. We
wish to learn about a particular unknown quantity $x$, of which we are
able to make direct or indirect noisy observations. For now
we say that $x$ is a member of a Hilbert
space $X$.

The quantity $x$ is mapped on to observable space by the observation
operator $\mathcal{G}:X \to\mathbb{R}^d$. We assume that the
observations, $\data$, are subject to Gaussian noise,
\begin{equation}\label{eqn:obs}
	\data = \mathcal{G}(x) + \varepsilon, \qquad \varepsilon \sim \mathcal{N}(0,\Sigma).
\end{equation}

@@

Then, according to Bayes' theorem, the posterior density that we
are interested in characterising is given, up to a constant of proportionality, by
\begin{equation*}
	\pi_Y \propto \exp \left ( -\Phi(x^*) \right )\pi_0(x^*),
\end{equation*}
where $\pi_0$ is the prior density on the quantity of interest $x^*$.

@@

In this work we use the resampler used within these particle filters. We
approximate the posterior {\red density $\pi$,} with a sample
of weighted particles, $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
the weights would be found by incorporating new observed data whereas
here we simply use importance weights. These weighted states can then be resampled, using the ETPF or otherwise,
into a {\red set of new equally-weighted states, such that
\begin{equation}\label{eqn:resampler}
	\sum\limits_{i=1}^M \! w_i\delta_{y_i}(\cdot)
	 \quad \xrightarrow{\text{ETPF}} \quad \sum\limits_{i=1}^M \!
	 \frac{1}{M} \delta_{x_i}(\cdot).
\end{equation}}

@@

We wish to sample states $x \in X$ from a posterior
probability distribution with density {\red $\pi$}. Since we have $M$ ensemble members, we
represent the current state of all of the Markov chains as a vector
$\X = [x_1,x_2,\ldots,x_M]^\top$. We are also given a transition kernel
$\nu(\cdot,\cdot)$, which might come from an MCMC method, for example
the random walk Metropolis-Hastings proposal density $\nu(\cdot,x) \sim
\mathcal{N}(x,\beta^2)$, where $\beta^2\in \mathbb{R}$ defines the
variance of the proposal.

@@

Care needs to be taken when choosing kernel(s) for the proposal distribution to
ensure that the overall mixture is absolutely continuous with respect to
the posterior distribution. {\red For example, in many inverse problems we
may be looking to find the value of certain physical parameters which
may be strictly positive, or even bounded on an interval. In
this case, proposal kernels $\nu$ should be picked which have the same
support as the target. We will see an example of an unknown parameter
being bounded in Section \ref{sec:mixture}, where
$\beta$-distributed proposals are made to ensure that the distribution
is supported on $[0,1]$.}

@@

\section{Consistency of ETAIS}\label{sec:consistency}
{\red As with all importance sampling schemes, we must have absolute
  continuity of target distribution with respect to the proposal
  distribution if we wish to achieve convergence. This can usually be
  ensured by picking the proposal kernels $\nu$ to be from the same
  distribution type as the prior distribution.}

@@

{\red
\section{A useful property of the ETAIS algorithm for multimodal
  distributions} }

@@

\subsection{Multimodal targets and the effect of resampler quality}
\label{sec:bimodal}
{\red
In this second example we investigate the behaviour of the ETAIS
algorithm when applied to a bimodal problem, given by a mixture of two
Gaussians,
\[\pi(x) = 0.2\pi_1(x) + 0.8\pi_2(x),\]
where $\pi_1$ is the density of a $\mathcal{N}\left (\begin{pmatrix} 1
    \\ 1 \end{pmatrix}, \begin{pmatrix} 0.1 & 0 
    \\ 0 & 0.1 \end{pmatrix} \right )$ random variable and $\pi_2$ is the
  density of a $\mathcal{N}\left (\begin{pmatrix} -5
    \\ -5 \end{pmatrix}, \begin{pmatrix} 2.75 & -2.25 
    \\ -2.25 & 2.75 \end{pmatrix} \right )$ random variable.}

@@

Figure \ref{kshdbf} shows the bimodal target density, and convergence
plots for the RWMH and ETAIS-RW with ETPF, AMR and bootstrap
resamplers. As expected, a higher quality resampler leads to better
proposal distributions, which in turn leads to greater stability and
more reliable convergence. However, this also shows that the AMR is a
good greedy approximation of the ETPF, and for a fraction of the cost
when the number of ensemble members is larger, as shown in Section
\ref{sec:AMR}. The RWMH algorithm fails to converge efficiently since chains only very rarely
make switches between the two modes, leading to very slow mixing. This
demonstrates the advantage that ensemble-based methods can have over
other methods when the target
is multimodal.

@@

\subsubsection{Implementation}
\label{sec:mixture_implementation}
{\red In order to quantify error, we note that} the probability density should
be evenly divided between the two modes. This is due to the symmetric
prior for $p$, and and the same priors being assigned to $\mu_1$ and
$\mu_2$, and also to $\sigma^2_1$ and $\sigma^2_2$. To decide which
mode a sample belongs to we define a plane which bisects the posterior
so that each point on this plane lies exactly halfway between the two
true solutions to the inverse problem i.e. the value of $\theta$ used
to generate the data, and also the $\theta$ obtained by a relabelling
of the parameters. Now that we can assign a sample to a particular
mode, we can calculate the density in each mode by summing the weights
associated to all samples in that mode,
\[
	\bar{w}_k = \sum\limits_{i=1}^N w_iI_{X_i \in \text{Mode $k$}}, \quad k = 1, 2,
\]
and the relative error in the amount of density in each mode is then
\begin{equation}\label{eqn:mode_prop}
	w_\text{error} = 2\left|\frac{\bar{w}_1}{\bar{w}_1+\bar{w}_2} - \frac{1}{2}\right|.
\end{equation}

Since the probability $p$ is constrained to lie in the interval
$[0,1]$, and the variances must be positive, it can be wasteful to use
Gaussian proposal distributions, {\red which will produce samples outside of
the support of the posterior. Moreover, the value of the variances of
the unknown distributions are strictly non-negative. The algorithm
will be most efficient if the proposal and posterior distributions are
mutually absolutely continuous. It is also useful to be able to pick
proposal densities for which the variance is easily scaled so that we
can tune them to optimise efficiency. Thus we pick the following
proposal distributions for the $p$, $\mu$ and $\sigma^2$ parameters, respectively;}
\[
	q_p \sim \text{Beta}(\delta^{-2}p, \delta^{-2}(1-p)), \quad q_{\mu_{1,2}} \sim \mathcal{N}(\mu_{1,2}, 4\delta^2) \quad \text{and} \quad q_{\sigma^2_{1,2}} \sim \text{Gamma}(\alpha^*, \beta^*),
\]
where $\alpha^* = \sigma^2_{1,2}\beta^*$, $\beta^* =
\sigma^2_{1,2}/2\delta^2$ and $\delta$ is a scaling parameter to be
tuned. This means that our proposal distributions will not be a
mixture of multivariate Gaussians, but independent mixtures of
univariate Beta, Gamma and Gaussian distributions.

@@

For this example, we observe noisily the position of a particle at ten
equally spaced points in the time interval $t \in (0, 1]$. The Lorenz
equations are solved with the chaotic parameters given above and the
initial condition $\mathbf{x}_0^1$. A time step of $h=1\times 10^{-3}$
was used to evolve the equations numerically with the explicit Euler
method. The noise added to each observation was taken from the
distribution {\red $\mathcal{N}(0, 0.1^2I)$.} Priors for the initial condition
coordinates were given by
\[
	x_0 \sim \mathcal{N}(-0.5, 0.03^2), \quad y_0 \sim\mathcal{N}(-0.5, 0.03^2) \quad \text{and} \quad z_0 \sim \mathcal{N}(15, 0.4^2).
\]
The posterior density function takes the form
\[
	\pi(\mathbf{x}_0|D) \propto \exp\left\{-\frac{1}{2}\|\mathcal{G}(\mathbf{x}_0) - D\|^2_\Sigma\right\}\mathcal{N}(\mathbf{x}_0; m, 0.03^2\text{I}),
\]
where $\mathcal{G}$ calculates the solution to the Lorenz equations, using the Euler method with a time step of $h=1\times 10^{-3}$ starting at the initial condition $\mathbf{x}_0$, and returns the position at $t = 0.1, ..., 1$. The mean of the prior distribution $m = [-0.5, -0.5, 15]$.

*********
\begin{figure}[htb]
\centering
\includegraphics[width=0.8\textwidth]{"figures/Lorenz_posterior"}
\caption{Posterior distribution for the initial condition of the particle with motion governed by the Lorenz equations as described in Section~\ref{sec:lorenz_target}.}
\label{fig:Lorenz_posterior}
\end{figure}
********* NEW FIGURE HERE

@@

For this example, we use an ensemble size of $M=1500$, and the AMR resampling algorithm. The optimal scaling parameters are calculated for both algorithms using test runs which are not included in the convergence cost calculations. Convergence graphs are produced using 32 repeats of each algorithm with each simulation producing one million samples.

@@

% The optimal scaling parameters for the MH and ETAIS algorithms are given in Table~\ref{table:Lorenz_opt_parameters}.

% \begin{table}[!htb]
%       \centering
%         \begin{tabular}{|l|r|r|}
% 	\hline
% 	Algorithm	& MH & ETAIS \\ \hline
% 	$\delta^*$	& 1.0e-1     & 1.3e-2 \\
% 	\hline
% 	\end{tabular}
% 	\vspace{1mm}
% 	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the ETAIS algorithm is optimised using the effective sample size.}
% 	\label{table:Lorenz_opt_parameters}
% \end{table}

**********
\begin{figure}[htb]
\centering
\includegraphics[width=0.8\textwidth]{"figures/Lorenz_convergence"}
\caption{Convergence of the sample mean to the posterior mean for the MH and ETAIS algorithms. Posterior distribution as described in Section~\ref{sec:lorenz_target}.}
\label{fig:Lorenz_convergence}
\end{figure}
****************** NEW FIGURE HERE

@@

{\red The implementations of the ETAIS that have been presented in this
  paper, have been run in serial, but we argue that, particularly for
  inverse problems with expensive likelihoods, for example in the case
  where this requires the approximation of the solution of a
  differential equation, that this method is an excellent candidate
  for parallelisation, with the likelihood evaluations for each
  particle being computed on a different processor. This said, even in
  parallel, the numerical results that we have presented demonstrate
  that even a serial implementation can outperform standard
  Metropolis-Hastings methods.}
